---
title: "EDA of diamonds dataset"
author: "Saurabh Mathur"
date: "4 April 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction and setup
The `diamonds` dataset contains the prices of 50,000 round cut diamonds.
It is provided with the `ggplot2` library.

### Format
It has the following variables - 
- *price*: price in US dollars ($326–$18,823)
- *carat*: weight of the diamond (0.2–5.01)
- *cut*: quality of the cut (Fair, Good, Very Good, Premium, Ideal)
- *color*: diamond colour, from J (worst) to D (best)
- *clarity*: a measurement of how clear the diamond is (I1 (worst), SI1, SI2, VS1, VS2, VVS1, VVS2, IF (best))
- *x*: length in mm (0–10.74)
- *y*: width in mm (0–58.9)
- *z*: depth in mm (0–31.8)
- *depth*: total depth percentage = z / mean(x, y) = 2 * z / (x + y) (43–79)
- *table*: width of top of diamond relative to widest point (43–95)

### Loading the dataset
Loading the `ggplot2` library loads our dataset as well.
```{r}
library(ggplot2)
```

```{r}
sum(is.na(diamonds))
```
As seen above there are no `NA` values in the dataset as it has been pre-cleaned.

## Structure
The dataset has the following columns - 
```{r}
colnames(diamonds)
```

Its dimensions are given by -
```{r}
dim(diamonds)
```

The complete summary of the dataset -
```{r}
summary(diamonds)
```

## Tables

### Cut v/s Color
```{r}
table(diamonds$cut, diamonds$color)
```

### Cut v/s Clarity
```{r}
table(diamonds$cut, diamonds$clarity)
```

### Color v/s Clarity
```{r}
table(diamonds$color, diamonds$clarity)

```



## Plots


### Price 
```{r}
hist(diamonds$price)
```

### Carat
```{r}
hist(diamonds$carat)
```

### carat v/s Price

#### Simple 
A plot of carat v/s price of diamonds using the built-in `plot` function is given as -

```{r}
plot(x = diamonds$carat, y = diamonds$price, xlab = "Carat", ylab = "Price, in USD", main = "Carat v/s Price of diamonds (USD)")
```

#### qplot
A better-looking plot of carat v/s price with cut represented by color using the function `qplot` from the `ggplot2` library loaded earlier
```{r}
qplot(data = diamonds, x = carat, y = price, xlab = "Carat", ylab = "Price, in USD", main = "Carat v/s Price of diamonds (USD)", color = cut)
```

A zoomed in version of the earlier plot (excluding some outliers)
```{r}
qplot(data = diamonds, x = carat, y = price, xlab = "Carat", ylab = "Price, in USD", main = "Carat v/s Price of diamonds (USD)", color = cut) + scale_x_continuous(lim = c(0, 2.5), breaks = seq(0, 2.5, 0.25)) + scale_y_continuous(lim = c(0, 18000), breaks = seq(0, 18000, 500))

```


## Regression, Carat v/s Price

### Splitting the data
In order to perform cross-validation on our regression model, it is necessary to split the data into two parts namely, the `train` (80 %) and the `test` (20 %) datasets.
```{r}
test.indices <- sample(1:nrow(diamonds), size = 0.2 * nrow(diamonds))
test <- diamonds[test.indices, ] 
train <- diamonds[-test.indices, ]

# Sanity check
nrow(test) + nrow(train) == nrow(diamonds)
```

### Training the regression model
```{r}
model <- glm(data = train, formula = price ~ exp(carat))
```

### Plotting the fit
```{r}
xpoints <- seq(0, 5, 0.1)
ypoints <- predict(model, list(carat = xpoints))
plt <- ggplot() + geom_point(data = diamonds, aes(x = carat, y = price), color = "gray") 
plt <- plt + xlab("Carat") + ylab("Price in USD") + ggtitle("Diamonds, regression model : price ~ exp(carat)") 
plt <- plt + geom_line(aes(x = xpoints, y = ypoints))
plt <- plt + scale_x_continuous(lim = c(0, 4.25), breaks = seq(0, 5, 0.25)) 
plt <- plt + scale_y_continuous(lim = c(0, 18823), breaks = seq(0, 18823, 500))
plt
```

### Testing
The root mean square error for the model is calculated as follows -
```{r}
price.prediction <- predict(model, data = test)
rms.error <- sqrt(sum((price.prediction - test$price) ^ 2) / nrow(test))
rms.error
```